Sketch Recognition using Particle Swarm Optimization Algorithm

### 2009-04-20 trunk/draft/System.tex
i went to dr. magdah but she did not look at the thesis
second commit

  The direction is calculated as the angle between two vectors as in equation \ref{eq:direction} where $\overrightarrow {p_i p_j }$ is the vector from point $pi$ to point $pj$ and $\|{\overrightarrow {p_i p_j }}\|$ is the normal of the  vector $\overrightarrow{p_i p_j }$. The curvature is estimated using more than one method the result of each curvatures estimation is reported in section \ref{sec:Curvatureestimationresults}. The best curvature estimation is used, the estimation define curvature as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.

@@

\subsubsection{Polygon Division Algorithm \textsl{AlgS1}}
\label{sec:PolygonDivisionAlgorithm}
The first algorithm, the algorithm tries to segment the stroke as a polygon. The final solution is a set of line segments that best define the input stroke. Given the inputs stroke let's define the set of points in the stroke as  $S = \left\{ {x_1 ,x_2  \ldots x_N } \right\}$ which is the set of consecutive points from start of stroke till the end. The arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point $x_i,x_{i+1} \cdots,x_j$. The line
$\overline{x_i x_j} $ is the straight line connecting point $x_i$ to point $x_j$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1} 
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation}
 where $M$ is the number of dominant points in this solution.  The error $ e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$ \cite{PolygonApproximationPSO}. Figure \ref{fig:DPSOERROR} shows a graphical representation of the error. The fitness is computed using equation \ref{eq:fitnessSwarm1} 
\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation}%\]
where $N$ is the number of points in the stroke, $D$ is the number of point in the solution that was previously labeled as $P_{pd}$  and $E$ is the computed error and $\varepsilon$ is the error threshold.  If the solution produced an error that exceed the error threshold $\varepsilon$ the fitness function assign a -ve value for the solution to express it's in feasibility, otherwise the inverse of the number of vertex produced is used to access the solution fitness in terms of minimum number of vertices'.  This fitness function optimize two goals the first is to move the solution into the feasible solution space with acceptable error bounds and second is to fly the particle to a new position which may result in a polygon with a fewer number of vertices'\cite{PolygonApproximationPSO}. 
\begin{figure}
	\centering
  \includegraphics[scale=0.8]{images/pso1.eps}			
	\caption{AlgS1 Error}% a) Possible dominant point b) Particle encoding  
	\label{fig:DPSOERROR}
\end{figure}
 %The error is chosen to in favor of larger than the threshold is given a -ve value to lower the value of solution otherwise the system will favor the lower number of vertices'.\\ %we will want to lower the number of vertices'. \\
%if we say that s
%Alg1:
%Alg2: 
\subsubsection{Curve Divisions Algorithm \textsl{AlgS2}}
\label{sec:PolygonDivisionAlgorithm}

@@

\subsubsection{Solution Refine procedures} 
%\cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but we used another enhancement method.
 After each loop in the swarm algorithm, each particle loops on the set of selected dominant points to enhance the solution. Each dominant point is checked to find if it was labeled before as a possible dominant (computed as in section \ref{sec:Preprocessing}), if not the point is moved to the nearest labeled point.
   After each loop of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedure. For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{sec:Preprocessing}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. After that the particles are tested to make sure that the distance between every two successive dominant point is larger than the constant $min_D$. If two points are nearer than $min_D$\footnote{Constant selected to be 10 pixels} then one of the points is removed.

\subsection{Choosing the best fit}
The two PSO algorithms produce two segmentation solutions. After that the system evaluates each solution to finally select the best segmentation. This evaluation is based on the area of both the segmentation and the input stroke. The segmentation error($E_{alg}$ in Equation \ref{eq:bestFit2})represents the sum of Square Root Error of the length from the point on the stroke to the estimated curve or line. In \textsl{AlgS2}, this error is the same as the computed error in Equation \ref{eq:ErrorSwarm1} but For the second algorithm \textsl{AlgS2} it is different than the computed error in Equation \ref{eq:errorSwarm2}. The additional computation is necessary to  compare the two segmentation fairly. 
\begin{equation}
\label{eq:bestFit1}
E_{area}=A_{st}-A_{sg}
 \end{equation}
 \begin{equation}
\label{eq:bestFit2}
E_{tot}=E_{area}+E_{alg}
 \end{equation}
 where $A_{st}$ is the area of the input stroke, $A_{sg}$ is area of computed segmentation, $E_{alg}$ is the Segmentation Error. The segmentation which cross ponds to the minimum $E_{tot}$ is the best segmentation.

@@

\section{ Feature Extraction}%%Feature Set
\label{sec:FeatureExtraction}

The system uses a composite set of features include Rubine feature set \cite{gestureexample12}, Zernike moments \cite{HeloiseBeautification}, ink density \cite{GeometryAndDomain102} and some structural information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. The following list gives details on all the features used in the system.

\begin{description}
 \item [Structural and geometrical Features (FS1)] Features defines the structure of the geometrical symbol.  
\begin{description}
	 \item[Segments] Number of segments in the symbol.
	 \item [Strokes] Number of strokes or partial strokes that created the symbols.  
		\item[Primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item [Curves] Number of curves or ellipses in the symbol. 
		\item [Lines] Number of lines in the symbol. 
		\item [Perpendicular lines] Number of perpendicular lines. 
		\item [Parallel lines] Number of parallel lines. 
		\item [Intersections] Number of intersection between lines and curves. 
		\item [T intersections] Number of T intersections. 
		\item [L intersections] Number of L intersections. 
		\item [X intersections] Number of X intersections.
		\item [Size Ratio] Ratio between width to height of the symbol.

@@

\end{description}
\item [Rubine Feature Set (FS2)] Features introduced by Rubine\cite{gestureexample12} for single stroke gestures. To compute the features we append all segments points into single path $ink_{path}$. The features is computed on the generated path $ink_{path}$.%then compute the features based on this path.  
\begin{enumerate}
	\item Cosine of starting angle.
	\item Sine of starting angle.
	\item Length of diagonal of bounding box. It gives an idea of the size of the bounding box).
	\item Angle of diagonal. It gives an idea of the shape of the bounding box (long, tall, square).
	\item Distance from start to end.  
	\item Total stroke length
	\item Change in Rotation( Arctan ). It gives the directional angle.
	\item Absolute rotation 
	\item Rotation squared 
	\item The maximum speed reached (squared) 
	\item Total time of stroke 
\end{enumerate}

\item[Statistical Features (FS3)]   
\begin{description}
	\item[Zernike moments] Zernike moments of order n\footnote{ We tried different values of n. Results show the analysis of each order.} \cite{HeloiseBeautification}. 
\end{description}
	\item[Composite Features (FS4)] Features that are composites of different geometrical and statistical symbol. 
	\begin{description}

	\item [Ink density] Compute the density of points inside its bounding box\cite{GeometryAndDomain102}.   
 	\item[Convex Hull Area] Area of convex hull with respect to area of bounding box of symbol.
	\item[Convex Hull Perimeter] Perimeter of convex hull with respect to total length of symbol.
		\item [Mean Centroidal radius] The Mean of the centroidal radius which is the distance from each point in the symbol to the center of gravity.

>>>>

\begin{figure}
	\centering
		\includegraphics{images/curvatureCal.jpg.eps}
	\caption{Direction Calculation}
	\label{fig:curvatureCal.jpg}
\end{figure}

  The direction is calculated as the angle between vectors $Q=$ and the x-xaxis as in equation \ref{eq:direction} where $\overrightarrow {p_i p_j }$ is the vector from point $pi$ to point $pj$ and $\|{\overrightarrow {p_i p_j }}\|$ is the normal of the  vector $\overrightarrow{p_i p_j }$. The curvature is estimated using more than one method the result of each curvatures estimation is reported in section \ref{sec:Curvatureestimationresults}. The best curvature estimation is used, the estimation define curvature as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.

@@

\subsubsection{Polygon Division Algorithm \textsl{AlgS1}}
\label{sec:PolygonDivisionAlgorithm}
The first algorithm, the algorithm tries to segment the stroke as a polygon. The final solution is a set of line segments that best define the input stroke. Given the inputs stroke let's define the set of points in the stroke as  $S = \left\{ {x_1 ,x_2  \ldots x_N } \right\}$ which is the set of consecutive points from start of stroke till the end. The arc $\widehat{x_ix_j}$ is defined as the consecutive set of points from point $x_i,x_{i+1} \cdots,x_j$. The line
$\overline{x_i x_j} $ is the straight line connecting point $x_i$ to point $x_j$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1} 
\begin{equation}
E=\sum\nolimits_{i = 0}^M e ( \widehat{x_ix_{i+1}},\overline{x_i x_{i+1}})
\label{eq:ErrorSwarm1}
\end{equation}
 where $M$ is the number of dominant points in this solution.  The error $ e ( \widehat{x_ix_j},\overline{x_i x_j})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_ix_j}$ to the line $\overline{x_i x_j}$ \cite{PolygonApproximationPSO}. Figure \ref{fig:DPSOERROR} shows a graphical representation of the error. The fitness is computed using equation \ref{eq:fitnessSwarm1} 
\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation}%\]
where $N$ is the number of points in the stroke, $D$ is the number of point in the solution that was previously labeled as $P_{pd}$  and $E$ is the computed error and $\varepsilon$ is the error threshold.  If the solution produced an error that exceed the error threshold $\varepsilon$ the fitness function assign a -ve value for the solution to express it's in feasibility, otherwise the inverse of the number of vertex produced is used to access the solution fitness in terms of minimum number of vertices'.  This fitness function optimize two goals the first is to move the solution into the feasible solution space with acceptable error bounds and second is to fly the particle to a new position which may result in a polygon with a fewer number of vertices'\cite{PolygonApproximationPSO}. 
\begin{figure}
	\centering
  \includegraphics[scale=0.8]{images/pso1.eps}			
	\caption{AlgS1 Error}% a) Possible dominant point b) Particle encoding  
	\label{fig:DPSOERROR}
\end{figure}
 %The error is chosen to in favor of larger than the threshold is given a -ve value to lower the value of solution otherwise the system will favor the lower number of vertices'.\\ %we will want to lower the number of vertices'. \\
%if we say that s
%Alg1:
%Alg2: 
\subsubsection{Hybird Curve Algorithm \textsl{AlgS2}}
\label{sec:PolygonDivisionAlgorithm}

@@

\subsubsection{Solution Refine procedures} 
%\cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but we used another enhancement method.
 After each loop in the swarm algorithm, each particle loops on the set of selected dominant points to enhance the solution. Each dominant point is checked to find if it was labeled before as a possible dominant (computed as in section \ref{sec:Preprocessing}), if not the point is moved to the nearest labeled point.
   After each loop of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedure. For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{sec:Preprocessing}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominant points $P_{pd}$. After that the particles are tested to make sure that the distance between every two successive dominant point is larger than the constant $min_D$. If two points are nearer than $min_D$\footnote{ 5\% of the total stroke length} then one of the points is removed.

\subsection{Choosing the Best Fit}
The two PSO algorithms produce two segmentation solutions. After that the system evaluates each solution to finally select the best segmentation. This evaluation is based on the area of both the segmentation and the input stroke. The segmentation error($E_{alg}$ in Equation \ref{eq:bestFit2})represents the sum of Square Root Error of the length from the point on the stroke to the estimated curve or line. In \textsl{AlgS2}, this error is the same as the computed error in Equation \ref{eq:ErrorSwarm1} but For the second algorithm \textsl{AlgS2} it is different than the computed error in Equation \ref{eq:errorSwarm2}. The additional computation is necessary to  compare the two segmentation fairly. 
\begin{equation}
\label{eq:bestFit1}
E_{area}=A_{st}-A_{sg}
 \end{equation}
 \begin{equation}
\label{eq:bestFit2}
E_{tot}=E_{area}+E_{alg}
 \end{equation}
 where $A_{st}$ is the area of the input stroke, $A_{sg}$ is area of computed segmentation, $E_{alg}$ is the Segmentation Error. The segmentation which cross ponds to the minimum $E_{tot}$ is the best segmentation.

@@

\section{Feature Extraction Methods}%%Feature Set
\label{sec:FeatureExtraction}

The system uses a composite set of features which include global shape properties like Rubine feature set \cite{gestureexample12},  ink density \cite{GeometryAndDomain102}, some appearance based properties like Zernike moments \cite{HeloiseBeautification}, and some stroke based structural information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. The following list gives details on all the features used in the system.

\begin{description}
 \item [Structural and geometrical Features (FS1)] Features defines the structure of the geometrical symbol.  
\begin{description}
	 \item[Segments] Number of segments in the symbol.
	 \item [Strokes] Number of strokes or partial strokes that created the symbols.  
		\item[Primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item [Curves] Number of curves or ellipses in the symbol. 
		\item [Lines] Number of lines in the symbol. 
		\item [Perpendicular lines] Number of perpendicular lines. 
		\item [Parallel lines] Number of parallel lines. 
		\item [Intersections] Number of intersection between lines and curves. 
		\item [T intersections] Number of T intersections. 
		\item [L intersections] Number of L intersections. 
		\item [X intersections] Number of X intersections.

@@

\end{description}
\item [Rubine Feature Set (FS2)] Features introduced by Rubine\cite{gestureexample12} for single stroke gestures. These features represent the global shape of the symbol drawn. To compute the features we append all segments points into single path $ink_{path}$. The features is computed on the generated path $ink_{path}$.%then compute the features based on this path.  
\begin{enumerate}
	\item Cosine of starting angle.
	\item Sine of starting angle.
	\item Length of diagonal of bounding box. It gives an idea of the size of the bounding box).
	\item Angle of diagonal. It gives an idea of the shape of the bounding box (long, tall, square).
	\item Distance from start to end.  
	\item Total stroke length
	\item Change in Rotation( Arctan ). It gives the directional angle.
	\item Absolute rotation 
	\item Rotation squared 
	\item The maximum speed reached (squared) 
	\item Total time of stroke 
\end{enumerate}

\item[Statistical Features (FS3)]   
\begin{description}
	\item[Zernike moments] Zernike moments of order n\footnote{ We tried different values of n. Results show the analysis of each order.} \cite{HeloiseBeautification}. 
\end{description}
	\item[Composite Features (FS4)] Features that are composites of different geometrical and statistical symbol. These features represent the global shape of the symbol drawn.
	\begin{description}
	\item [Size Ratio] Ratio between width to height of the symbol.
	\item [Ink density] Compute the density of points inside its bounding box\cite{GeometryAndDomain102}.   
 	\item[Convex Hull Area] Area of convex hull with respect to area of bounding box of symbol.
	\item[Convex Hull Perimeter] Perimeter of convex hull with respect to total length of symbol.
		\item [Mean Centroidal radius] The Mean of the centroidal radius which is the distance from each point in the symbol to the center of gravity.
